Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2019-02-23 accessPEP.Rmd
with comparison of burden across scenario analysis and scale of estimation

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
---

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())

## Libraries
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
library(stringr)
library(ISOweek)
library(raster)
library(sp)
library(MASS)
library(reshape2)
source("R/utils.R")

@@

### Bite patient data
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted as frequently as monthly to annually by clinics. Of the 31 existing ARMC, `r length(no_data) - 2` submitted fewer than 10 forms over the four years. Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis. Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A). These data include the administrative district of the bite patient's address and the date of reporting. We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below district).

@@

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1). Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC. We excluded any districts from catchments for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).

@@

As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2 shows the time series of form submission for each clinic and periods of time excluded by the 15 day threshold). Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold. To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

@@

Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data. We also compared this method to assuming that 20% of total bites at the district level were contacts.

@@

### Model of reported bites as a function of travel time
While the national bite patient data is available at the district level, travel times and therefore reporting, likely vary significantly at the sub-district level. In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level. Incidence at the commune level is then a function of travel times to the closest ARMC. Specifically, we modeled bites as follows:

$$ 
  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
$$
where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$. We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.

As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:

$$ 
\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
$$

where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$. We only included communes which were designated to be within the catchment for the clinic. We compared our estimates of $\beta_t$ (i.e. the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.

Finally, we compared these models to a model of bites at the district level as the function of travel times averaged to the district level:
$$ 
  \mu_{d} = exp(\beta_{t}T_d + \beta_0)\times pop_d 
$$
where $pop_d$, $T_d$, and $mu_d$ are respectively human population size, weighted travel times, and mean number of bites in district $d$. As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models, as a test of whether the observed patterns could be predicted by bite incidence scaling with population size. We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.

@@

### Strategically expanding PEP access (TO DO)
Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional ARMC. Starting with the current locations, we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country. We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic in each iteration sequentially. We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added (taking the min/max approach).

### Sensitivity analyses for burden estimates and scenario analyses (TO DO)

@@

@@

When we compare our burden estimates at the district vs. the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden

@@

### Strengths and Limitations
- Not accounting for clinic functioning
- Not explicit data on reporting or rabies incidence!
- Other factors that drive reporting
- While we use a signficiant amount of assumptions, these are the same set of assumptions that underlie current rabies estimates, we're just improving them a bit... and ours are a likely a little bit more in line with reality @ a sub-national level?

@@

>>>>

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
---

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## Libraries
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
library(stringr)
library(ISOweek)
library(raster)
library(sp)
library(MASS)
library(reshape2)
source("R/utils.R")

@@

@@

### Bite patient data
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted as frequently as monthly to annually by clinics. Of the 31 existing ARMC, `r length(no_data) - 2` submitted fewer than 10 forms over the four years. Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis. Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A). These data include details of the administrative district of the bite patient's address and the date of reporting. We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1). Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC. We excluded any catchment areas for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).

@@

As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2). Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold. To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

@@

Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data. We also compared this method to assuming that 20% of total bites at the district level were contacts.

@@

### Model of reported bites as a function of travel time
We modeled bites as a function of travel times as follows:
$$ 
  \mu = exp(\beta_{t}T + \beta_0)P 
$$
where $mu$ is the expected number of bites given travel times and population for a given location. In this way, we can model the impact of travel time on incidence as 
$exp(\beta_{t}T + \beta_0)$ gives an estimate of incidence and multiplying by the population size gives the expected number of bites, so that we can fit this model to the number of bites in each location.

We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.

We fit this to our available data in three ways:
1) the commune data available at the moramanga district, where we fit to $mu_j$
2) The district data available nationally
3) To compare estimates from 1 directly to the national data, the district data but where $mu_d$ is the sum of the travel time effect at the commune level, i.e.:

$$ 
  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
$$

 As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models (replacing $Tj$ with log-transformed population in $j$ to avoid squared terms), as a test of whether the observed patterns could be predicted by bite incidence scaling with population size alone. We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.

@@

### Strategically expanding PEP access
Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional ARMC. Starting with the current locations, we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country. We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added.

### Sensitivity analyses for burden estimates 
- R (increasing or decreasing with pop) (min + max)
- P-rabid min and max
- Probability of death min and max
- Look at change in raw # of deaths and also change in burden by scenarios

@@

```{r scale example}
## Comparing scales
## Function for reporting
get.reporting_burden <- function(pop, ttimes, B_ttimes, B_0, hdr = 25, incidence = 0.01, 
                          exp_rate = 0.39, p_rabid = 0.6, rho_max = 0.98, p_death = 0.16, 
                          type = "reporting") {
  bites_100k <- exp(B_ttimes*ttimes + B_0)*pop
  R_100k <- pop/hdr*incidence*exp_rate
  if (type == "reporting"){
    rho <- bites_100k*(ifelse(bites_100k*p_rabid > R_100k, rho_max*R_100k/bites_100k, p_rabid))/R_100k
    return(rho)
  } else {
    unreported <- R_100k - bites_100k*(ifelse(bites_100k*p_rabid > R_100k, 
                                              rho_max*R_100k/bites_100k, p_rabid))
    deaths <- 0.16*unreported
    return(deaths)
  }
}

comm_data$reporting_min <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
                                            hdr = 5, p_rabid = 0.2)

comm_data$reporting_max <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
                                            hdr = 25, p_rabid = 0.6)
comm_data %>%
  group_by(district) %>%
  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
  summarize(reporting_min = sum(reporting_min*prop_pop), 
            reporting_max = sum(reporting_max*prop_pop)) %>%
  mutate(reporting_med = (reporting_min + reporting_max)/2,
         type = "summed from commune") -> summed_rep_dist

mada_district@data %>%
  mutate(reporting_min = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
                                              p_rabid = 0.2), 
         reporting_max = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
                                              p_rabid = 0.6),
         reporting_med = (reporting_min + reporting_max)/2,
         type = "district level") %>%
  select(district, reporting_min, reporting_max, reporting_med, type) -> rep_comp

rep_comp <- rbind(rep_comp, summed_rep_dist)
rep_comp$ttime <- mada_district$ttimes_weighted[match(rep_comp$district, mada_district$district)]

ggplot(rep_comp, aes(x = reporting_med, y = reorder(district, ttime), 
                              color = type)) +
  ggstance::geom_pointrangeh(aes(xmin = reporting_min, xmax = reporting_max), fatten = 2)

bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
R_100k_min <- pop_plot/25*0.01*0.39
R_100k_max <- pop_plot/5*0.01*0.39

plot(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
     type = "l", bty = "n",
     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
     col = "darkred")
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
     col = "red", lty = 2)
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
     col = "red", lty = 2)
polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
        border = NA, col = alpha("red", 0.2))
polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
        border = NA, col = alpha("red", 0.6))

## Mora
comm_data %>%
  filter(district == "Moramanga") -> comm_level
bites_100k <- exp(B_ttimes_mada*comm_level$ttimes_weighted + B_0_mada)*comm_level$pop
R_100k_min <- comm_level$pop/25*0.01*0.39
R_100k_max <- comm_level$pop/5*0.01*0.39

## district val for Moramanga
## need to merge names
exps_dist %>%
  filter(dist_name == "Moramanga") %>%
  mutate(rep_max = get.reporting_burden(pop = pop, ttimes = ttimes_weighted_dist, 
                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
                                              p_rabid = 0.6),
         rep_min = get.reporting_burden(pop = pop, ttimes = ttimes_weighted_dist, 
                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
                                              p_rabid = 0.2)) -> dist_level
dist_level %$% points(ttimes_weighted_dist/60, rep_max, col = "red", pch = 20)
dist_level %$% points(ttimes_weighted_dist/60, rep_min, col = "red", pch = 20)
dist_level %$% abline(v = ttimes_weighted_dist/60, col = "red", lty = 2)

@@

When we compare our burden estimates at the district vs. the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden.

### Expanding access to PEP

```{r scenario analysis}
## new ctar
expanded <- read.csv("output/incremental_ARMC_nofilter.csv")
dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)

dist_burden_min <- apply(dist_ttimes, 2, function(x) get.reporting_burden(pop = mada_district$pop2015adj, ttimes = x, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, p_rabid = 0.6, type = "burden"))
dist_burden_max <- apply(dist_ttimes, 2, function(x) get.reporting_burden(pop = mada_district$pop2015adj, ttimes = x, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, p_rabid = 0.2, type = "burden"))

comm_burden_min <- apply(comm_ttimes, 2, function(x) get.reporting_burden(pop = mada_communes$MDG__201, ttimes = x, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 25, p_rabid = 0.6, type = "burden"))
comm_burden_max <- apply(comm_ttimes, 2, function(x) get.reporting_burden(pop = mada_communes$MDG__201, ttimes = x, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 5, p_rabid = 0.2, type = "burden"))

## Decrease in burden over times
plot(colSums(comm_burden_min, na.rm = TRUE)/sum(comm_burden_min[,1], na.rm = TRUE), type = "l", col = "blue", bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", 
     ylab = "Proportion of deaths compared to baseline ")
lines(colSums(dist_burden_min, na.rm = TRUE)/sum(dist_burden_min[, 1], na.rm = TRUE),
      col = "purple")
lines(colSums(comm_burden_max, na.rm = TRUE)/sum(comm_burden_max[, 1], na.rm = TRUE), 
     type = "l", col = "blue", 
     bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", ylab = "Number of deaths")
lines(colSums(dist_burden_max, na.rm = TRUE)/sum(dist_burden_max[, 1], na.rm = TRUE), col = "purple")

@@

### Strengths and Limitations

@@

## Acknowledgements

## Supplementary Figures

#### S1
#### S2
#### S3
#### S4
#### S5
#### S6
